Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-12 talk/iwtc11/paper.tex
fix some formatting stuff, start to import some papers that we will need into the bib file

\usepackage{amsmath}
\usepackage{setspace}
\usepackage{listings}

@@

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Title Text}
\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

@@

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

Using these classes to implement arithmetic shows the basic problem of a
dynamic language implementation. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

\subsection{Allocation removal}
By using escape analysis it is possible to identify objects that are
allocated within the loop but never escapes it. That is the object are
short lived and no references to them exists outside the loop. This
is performed by processing the operation from top to bottom and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get} and \lstinline{set}) on the removed objects are also
removed and the optimizer needs to keep track of the value of all
attributes of the object.

Consider again the original unoptimized trace of
Figure~\label{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that was not yet
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments for the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all it's attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the second
operation, $K$, are replaced by inlining $\hat J$, 
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
            # inside BoxedInteger.__init__
jump($l_1$, $p_{0}$, $i_{4}$)

$l_1$($p_{0}$, $i_{4}$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{7}$ = get($p_{0}$, intval)
        $i_{8}$ = int_add($i_{4}$, $i_{7}$)
            # inside BoxedInteger.__init__
jump($l_1$, $p_{0}$, $i_8$)
\end{lstlisting}

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the first iteration. This
explosion can't be repeated when constructing the arguments of the
jump of the second iteration as it has to mach the first. This means
the objects that was passed as pointers (non virtuals) from the first
iteration to the second also has to be passed as pointers from the
second iteration to the third. If one of these objects are virtual
at the end of the second iteration they need to be allocated right
before the jump.

@@

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

@@

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

>>>>

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

\usepackage{listings}

\usepackage[T1]{fontenc}
\usepackage{setspace}
\usepackage{listings}
\usepackage{beramono}

\definecolor{gray}{rgb}{0.3,0.3,0.3}

\lstset{
  basicstyle=\setstretch{1.05}\ttfamily\footnotesize,
  language=Python,
  keywordstyle=\bfseries,
  stringstyle=\color{blue},
  commentstyle=\color{gray}\textit,
  fancyvrb=true,
  showstringspaces=false,
  %keywords={def,while,if,elif,return,class,get,set,new,guard_class}
  numberstyle = \tiny,
  numbersep = -20pt,
}

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}
\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}

\newcommand\ie{i.e.,\xspace}
\newcommand\eg{e.g.,\xspace}
\newcommand{\etal}{\emph{et al.}\xspace}

\normalem

\let\oldcite=\cite

\renewcommand\cite[1]{\ifthenelse{\equal{#1}{XXX}}{[citation~needed]}{\oldcite{#1}}}

@@

\conferenceinfo{IWTC '11}{XXX} 
\copyrightyear{2011} 
\copyrightdata{[to be supplied]}

\titlebanner{draft}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Loop Invariant Code Motion in PyPy's Tracing JIT}
%\subtitle{Subtitle Text, if any}

\authorinfo{Hakan Ardo XXX}
           {Affiliation1}
           {hakan@debian.org}
\authorinfo{Carl Friedrich Bolz}
           {Heinrich-Heine-Universität Düsseldorf}
           {cfbolz@gmx.de}

@@

\category{D.3.4}{Programming Languages}{Processors}[code generation,
incremental compilers, interpreters, run-time environments]

\terms
Languages, Performance, Experimentation

\keywords{Tracing JIT, Optimization, Loop-Invariant Code Motion}

@@

Using these classes to implement arithmetic shows the basic problem of many
dynamic language implementations. All the numbers are instances of either
\lstinline{BoxedInteger} or \lstinline{BoxedFloat}, therefore they consume space on the
heap. Performing many arithmetic operations produces lots of garbage quickly,
putting pressure on the garbage collector. Using double dispatching to
implement the numeric tower needs two method calls per arithmetic operation,
which is costly due to the method dispatch.

@@

\subsection{Virtualization}
Using escape analysis we can XXX

Let $\tilde J$ be all variables in $J$ not representing virtuals (in the
same order). Extend it with all non virtual fields, $H_i$, of the
removed virtuals,
\begin{equation}
  \hat J = \left(\tilde J_1, \tilde J_2, \cdots, \tilde J_{|\tilde J|}, 
                 H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
and let
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = int_add($i_{2}$, $i_{3}$)
jump($l_1$, $p_{0}$, $i_3$, $i_4$)

$l_1$($p_{0}$, $p_{5}$, $i_3$, $i_4$):
# inside f: y = y.add(step)
    # inside BoxedInteger.add
        # inside BoxedInteger.add__int
        $i_{8}$ = int_add($i_{4}$, $i_{3}$)
jump($l_1$, $p_{0}$, $i_3$, $i_8$)
\end{lstlisting}

And we're down to a single integer addition!

@@

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

@@

\bibliographystyle{abbrvnat}
\bibliography{paper}
